\pkg{digiRhythm}: an \proglang{R} package for rhythmicity assessment using the degree of functional coupling

### 2023-06-14 jstatsoft/article.Rnw
Discussion done

\section{Discussion} \label{sec:summary}
% Scope of DFC application
The \pkg{digiRhythm} package presents a novel, user-friendly, and accessible tool for researchers in the field of rhythmicity, animal behavior, and beyond. The package offers a versatile approach to assessing rhythmicity based on digital activity records. By leveraging the power of \proglang{R}, it brings robust statistical capabilities and intuitive visualization features into the hands of researchers. While the current implementation of the \pkg{digiRhythm} package focuses on rhythmicity in farm animals, the underlying principles and methodology can be extended to other organisms, including humans. This opens up exciting avenues for future research in various domains. The unique utilization of harmonic frequencies related to the 24-hour day in the package provides a novel approach for rhythmicity analysis. The package is designed with flexibility and expansion in mind. The current version allows for the computation of different metrics like DFC, DI, and visualization tools such as actograms. Still, the modular structure of the package enables the incorporation of additional techniques and tools in the future. As the \pkg{digiRhythm} package continues to develop and evolve, we expect it to become an off-the-shelf tool for researchers interested in studying rhythmicity patterns and their implications. Rhythmicity can be calculated in the \pkg{digiRhythm} package from any type of regularly measured time series data, if the data are prepared in a suitable format. The suitability of the format can be assessed with the  \fct{is\_dgm\_friendly} function. At the time of writing this paper, the current version (v1.2) of the library does not provide the capability to compute rhythmicity using techniques such as cosinor analysis (\cite{refinetti2016}). However, there is potential for future expansion of the library's functionality to incorporate additional techniques.

To date, this is a computation of activity patterns, resulting in the DFC. This measure has been correlated with physiological functions in previous studies (\cite{Berger, Saraout}). The computation of rhythmicity allows to gain further insight into an individual's harmonic patterns. Organisms in a good physiological state will present harmonic locomotion patterns, whereas this rhythmicity desynchronizes as a result of physiological events. In the past, these rhythmic patterns have been discussed with regard to the psychological state of humans, where  bipolar disorders or depression have been associated with the desynchronization of rhythmic patterns (\cite{sabet2021the}). However, there is no evidence to whether these activity patterns are intrinsically or extrinsically motivated. Such extrinsic factors could affect rhythmic activity patterns and blur the organisms intrinsic rhythmicity. To date, the package provides the option to exclude extrinsically structured time spans from the analysis. However, in order to increase the understanding of rhythmicity  future studies should investigate how these extrinsic factors affect rhythmicity calculations.

@@

The current package has been developed for farm animals, but the principles of rhythmicity apply for any organism, including humans. Therefore we highly recommend future research to evaluate the \pkg{digiRhythm} package in other species. A previous review highlighted the potential of such rhythmicity computations in humans \citep{dijk2020novel}. Thus, it would be interesting to implement the \pkg{digiRhythm} package in this domain.

>>>>

\section{Discussion} \label{sec:summary}
% Scope of DFC application
The \pkg{digiRhythm} package presents a novel, user-friendly, and accessible tool for researchers in the field of rhythmicity, animal behavior, and beyond. The package offers a versatile approach to assessing rhythmicity based on digital activity records. By leveraging the power of \proglang{R}, it brings robust statistical capabilities and intuitive visualization features into the hands of researchers. While the current implementation of the \pkg{digiRhythm} package focuses on rhythmicity in farm animals, the underlying principles and methodology can be extended to other organisms, including humans. This opens up new opportunities for future research in various domains. The unique utilization of harmonic frequencies related to the 24-hour day in the package provides a novel approach for rhythmicity analysis. The package is designed with flexibility and expansion in mind. The current version allows for the computation of different metrics, such as DFC, DI, and visualization tools such as actograms. Still, the modular structure of the package enables the incorporation of additional techniques and tools in the future. As the \pkg{digiRhythm} package continues to develop and evolve, we expect it to become an off-the-shelf tool for researchers interested in studying rhythmicity patterns and their implications. Rhythmicity can be calculated in the \pkg{digiRhythm} package from any type of regularly measured time series data, if the data are prepared in a suitable format. The suitability of the format can be assessed with the  \fct{is\_dgm\_friendly} function. At the time of writing this paper, the current version (v1.2) of the library does not provide the capability to compute rhythmicity using techniques such as cosinor analysis (\cite{refinetti2016}). However, there is potential for future expansion of the library's functionality to incorporate additional techniques.

The package calculates the DFC, by computing harmonic frequencies of activity patterns. This measure has been correlated with physiological functions in previous studies (\cite{Berger2003, sarout2018assessment}). The computation of rhythmicity allows to gain further insight into an individual's harmonic patterns (\cite{trigonum1993circadian}). Organisms in a good physiological state will present harmonic locomotion patterns, whereas this rhythmicity desynchronizes as a result of physiological events. \cite{rufener2018finding} found that individual laying hens followed a consistent individual pattern, but this pattern would differ between individuals within the flock. (\cite{opperhuizen2016feeding} further showed that altered feeding schedules during resting phases cause changes in physiology, as well as desynchronization of liver and muscle rhythms in rats. In the past, these rhythmic patterns have been discussed with regard to the psychological state of humans, where  bipolar disorders or depression, as well as delirium development have been associated with the desynchronization of rhythmic patterns (\cite{sabet2021the, ANGELESCASTELLANOS2016285}). However, the effect of intrinsic or extrinsic motivation of these activity patterns is not yet fully understood. Such extrinsic factors could affect rhythmic activity patterns and blur the organisms intrinsic rhythmicity. To date, the package provides the option to exclude extrinsically structured time spans from the analysis. However, in order to increase the understanding of rhythmicity  future studies should investigate how these extrinsic factors affect rhythmicity calculations.

@@

The current package has been developed for farm animals, but the principles of rhythmicity apply for any organism, including humans. Therefore we highly recommend future research to evaluate the \pkg{digiRhythm} package in other species. A previous review highlighted the potential of such rhythmicity computations in humans \citep{dijk2020novel}. \citep{sani2015daily} were able to distinguish differences in 24 hrs activity patterns between humans in different sociocultural backrounds. Thus, it would be interesting to implement the \pkg{digiRhythm} package in this domain.
